---
title: "Can median household income and access to a vehicle be predictive of supermarket presence in Buffalo, NY?"
author: Eric Hughes
---

# Introduction 

Access to healthy food is not equal across all regions. Many factors influence food access such as income (Walker et al., 2010), vehicle access (Crowe et al., 2018), and proximity to quality grocery retailers (Moore & Diez Roux, 2006). The most basic measure of food access is proximity to supermarkets because supermarkets have the resources to carry a wide range of produce and have affordable prices (Moore & Diez Roux, 2006; Sallis et al., 1986), making their presence a good measure of local food security. Using the Moore & Diez (2006) definition of a supermarket, only grocery stores with 50 or more employees are considered a supermarket.

This project aims to look at Buffalo resident's proximity to supermarkets based off their median household income and access to a vehicle. To do this, two analysis will be run. For the first, Buffalo residents will be grouped into four different income brackets based on their median household income per census tract. For the second, Buffalo residents will be grouped into four brackets based on proportions of households within each census tract without access to a personal vehicle. Then, the distance to the nearest supermarket for each census tract centroid will be calculated. Each census tract will be joined to the median household income and vehicle access data sets and averages will be calculated for each bracket. Results will be analyzed and discussed in the Results and Conclusions sections. 

# Materials and methods

## Data sources

- US Census Bureau will provide census tract data of median household income, as well as demographic data, for Buffalo

- Open Data Buffalo for Neighborhood Boundary Shape file

- ESRI Business Analyst will provide all supermarket data

## Packages
```{r, message=F, warning=F}
library(sf)
library(tidyverse)
library(ggplot2)
library(sp)
library(ggmap)
library(dodgr)
library(osmdata)
library(classInt)
library(knitr)
library(geodist)
library(kableExtra)
library(stringi)
```


## Data cleaning
###Import the two data sets, filter for needed fields, and rename fields logically

```{r, message=F, warning=F}

income_csv <- read.csv(file = "median_income_data_2020.csv",
                              skip = 2, header = F)
income_data <- income_csv[c(1,2,323)]
colnames(income_data) <- c("geoid", "census.tract", "median.household.income")
income_data$geoid = as.character(income_data$geoid)

vehicle_access_csv <- read.csv(file = "vehicle_access.csv")
vehicle_access_data <- vehicle_access_csv[c(1,2,6)]
colnames(vehicle_access_data) <- c("geoid","census.tract","prop.wo.va")
```
###Create Buffalo border
```{r, message=F, warning=F}

buffalo_border <- st_read("buffalo_boundary") %>%
  st_transform(4326)
buffalo_sp <- as_Spatial(st_transform(buffalo_border, 4326))
buffalo_bbox <- buffalo_sp@bbox

basemap <- get_stamenmap(
  bbox = buffalo_bbox,
  zoom = 13,
  maptype = "toner-lite")
```

###Import census tracts and join with median income and vehicle access data. Then convert data types of specific fields for future analysis

```{r, message=F, warning=F}

census_tracts <- st_read("buffalo_census_tracts") 

income_shape <- census_tracts %>% left_join(income_data, by = c("GEOID" = "geoid"))
income_shape$INTPTLAT = as.numeric(income_shape$INTPTLAT)
income_shape$INTPTLON = as.numeric(income_shape$INTPTLON)
income_shape$median.household.income = as.integer(income_shape$median.household.income)

vehicle_shape <- census_tracts %>% left_join(vehicle_access_data, by = c("GEOID" = "geoid"))
vehicle_shape$INTPTLAT = as.numeric(vehicle_shape$INTPTLAT)
vehicle_shape$INTPTLON = as.numeric(vehicle_shape$INTPTLON)
vehicle_shape$prop.wo.va = as.double(vehicle_shape$prop.wo.va)

```

###Project data and create centroid for each census tract

```{r, message=F, warning=F}

income_proj <- income_shape %>% 
  st_transform(4326) 

vehicle_proj <- vehicle_shape %>% 
  st_transform(4326) 

st_cent <- st_centroid(income_proj)

```

##Create Buffalo supermarkets variable##
###Create a 0.5-mile border around Buffalo and select all supermarkets within this newly created buffer zone

```{r, message=F, warning=F}
border_poly <- buffalo_border %>%
  st_cast("POLYGON")
buffer_buffalo <- st_buffer(border_poly, dist = 804.672)

buffalo_supermarkets <- st_read("erie_county_supermarkets")%>%
  st_crop(buffer_buffalo)
```

###Create X and Y columns in buffalo_supermarkets and centroid data
```{r, message=F, warning=F}
coords = st_coordinates(buffalo_supermarkets)
xlon = coords[,1]
ylat = coords[,2]
bf_supermarkets <- buffalo_supermarkets %>%
  mutate(xlon,ylat) 

sf_coords <- st_as_sf(st_cent) 
cen_coords <- st_coordinates(sf_coords)
xlon = cen_coords[,1]
ylat = cen_coords[,2]

centroids <- sf_coords %>%
  mutate(xlon,ylat) %>%
  st_transform(4326)
centroids$id <- 1:max(nrow(centroids))
```

## Find distance from centroid to nearest supermarket
# Define Buffalo coordinate variable
```{r, message=F, warning=F}
xy_coords <- rbind (
  c (-78.79514, 42.96641),
  c (-78.91252, 42.82610))
xy <- data.frame (lon = xy_coords [, 1], lat = xy_coords [, 2])
```

###Calculate distances from centroids to supermarkets
```{r, message=F, warning=F}

buffalo_network <- dodgr_streetnet (pts = xy, expand = 0.2, quiet = FALSE)
graph <- weight_streetnet (buffalo_network, wt_profile = "motorcar")
graph <- graph [which (graph$component == 1), ]
from_x <- centroids$INTPTLON
from_y <- centroids$INTPTLAT
to_x <- bf_supermarkets$xlon
to_y <- bf_supermarkets$ylat
d <- dodgr_dists(graph = graph, from = cbind (from_x, from_y), to = cbind (to_x, to_y))
df <- apply(d, 1, FUN=min, na.rm=TRUE) %>% as.data.frame()
b <- st_as_sf(data.frame(centroids, df))
names(b)[20] <- "dist"
```

##Create income brackets##

```{r, message=F, warning=F}
b_df <- as_data_frame(b)

#median income brackets
less_than_20k <- income_proj %>%
  filter(median.household.income <= 20000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_40k <- income_proj %>%
  filter(median.household.income > 20000, median.household.income <= 40000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_60k <- income_proj %>%
  filter(median.household.income > 40000, median.household.income <= 60000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

greater_than_or_equal_60k <- income_proj %>%
  filter(median.household.income >= 60000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

#Vehicle access brackets
less_than_0.2 <- vehicle_proj %>%
  filter(prop.wo.va <= 0.2) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_0.4 <- vehicle_proj %>%
  filter(prop.wo.va > 0.2, prop.wo.va <= 0.4) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_0.6 <- vehicle_proj %>%
  filter(prop.wo.va > 0.4, prop.wo.va <= 0.6) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

greater_than_0.6 <- vehicle_proj %>%
  filter(prop.wo.va > 0.6) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))
```

##Calculate average distance for each bracket in meters##
```{r, message=F, warning=F}

ave_dist_less_than_20k <- mean(less_than_20k$dist) #1813.465
ave_dist_less_than_40k <- mean(less_than_40k$dist) #1926.935
ave_dist_less_than_60k <- mean(less_than_60k$dist) #1830.889
ave_dist_greater_than_60k <- mean(greater_than_or_equal_60k$dist) #1648.723

ave_dist_less_than_0.2 <- mean(less_than_0.2$dist) #1661.945
ave_dist_less_than_0.4 <- mean(less_than_0.4$dist) #1844.538
ave_dist_less_than_0.6 <- mean(less_than_0.6$dist) #1957.577
ave_dist_greater_than_0.6 <- mean(greater_than_0.6$dist) #2438.269
```

# Results

## Plots for median household income and vehicle access

The plot for median household income shows where supermarkets are in relation to each census tract and each census tract's income bracket. 

```{r, message=F, warning=F}

income_supermarket <- ggmap(basemap) +
  geom_sf(data = income_proj, aes(x = INTPTLON, y = INTPTLAT,  fill = median.household.income), 
          inherit.aes = F)+
  scale_fill_gradient(low = "yellow", high = "red") +
  geom_point(data = bf_supermarkets, aes(x = xlon, y = ylat, color = "Supermarket"), size = 3) +
  scale_color_manual(values = "green", name = "Supermarkets") +
  labs(title="All supermarkets within 0.5-miles of Buffalo, NY (2020)",
       subtitle="Median income data",
       caption="Median household income data from US Census Bureau,
       supermarket data from Data Axle",
       x = "Longitude",
       y = "Latitude") +
  theme(legend.direction = "vertical", axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1))+
  guides(fill=guide_legend(title="Median Household Income ($)", order = 1)) 
income_supermarket  

```

The plot for vehicle access data shows where supermarkets are in relation to each census tract and each census tract's vehicle access bracket. 
```{r, message=F, warning=F}

vehicle_supermarket <- ggmap(basemap) +
  geom_sf(data = vehicle_proj, aes(x = INTPTLON, y = INTPTLAT,  fill = prop.wo.va), 
          inherit.aes = F)+
  scale_fill_gradient(low = "#f1eef6", high = "#ce1256") +
  geom_point(data = bf_supermarkets, aes(x = xlon, y = ylat, color = "Supermarket"), size = 3) +
  scale_color_manual(values = "green", name = "Supermarkets") +
  labs(title="All supermarkets within 0.5-miles of Buffalo, NY (2020)",
       subtitle="Proportion of households without vehicle access",
       caption="Vehicle access data from US Census Bureau,
       supermarket data from Data Axle",
       x = "Longitude",
       y = "Latitude") +
  theme(legend.direction = "vertical", axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1))+
  guides(fill=guide_legend(title="Proportion of households 
without access to a vehicle")) 
vehicle_supermarket

```

## Summary tables

The median household income data table shows that there was not an obvious pattern of supermarket distribution based upon median household income. The highest median income bracket did on average have the shortest distance to a supermarket but the other three brackets had similar distances with the bracket for "Between 20K and 40K" with the furthest distance.
```{r, message=F, warning=F}

income_bracket <- data.frame(matrix(ncol = 1, nrow = 4))
colnames(income_bracket) = "Income bracket ($)"
income_bracket[1,1] = "20k or less"
income_bracket[2,1] = "Between 20K and 40k"
income_bracket[3,1] = "Between 40k and 60k"
income_bracket[4,1] = "More than 60k"

income_dists <- data.frame(matrix(ncol = 1, nrow = 4))
colnames(income_dists) = "Average distance to nearest supermarket (miles)"
income_dists[1,1] = round(ave_dist_less_than_20k / 1609.344, digits = 2)
income_dists[2,1] = round(ave_dist_less_than_40k / 1609.344, digits = 2)
income_dists[3,1] = round(ave_dist_less_than_60k / 1609.344, digits = 2)
income_dists[4,1] = round(ave_dist_greater_than_60k / 1609.344, digits = 2)

income_table_combine <- cbind(income_bracket,income_dists)

income_table <- kableExtra::kable(income_table_combine) 

kable_styling(income_table, "striped", position = "left", font_size = 12, html_font = "arial")

```

##The vehicle access data table reveals a pattern where people with less access to a vehicle have further distances to travel to reach a supermarket.
```{r, message=F, warning=F}

vehicle_bracket <- data.frame(matrix(ncol = 1, nrow = 4))
colnames(vehicle_bracket) = "Vehicle access bracket"
vehicle_bracket[1,1] = "Less than 20% of households don't have access to a vehicle"
vehicle_bracket[2,1] = "Between 20% and 40% of households don't have access to a vehicle"
vehicle_bracket[3,1] = "Between 40% and 60% of households don't have access to a vehicle"
vehicle_bracket[4,1] = "More than 60% of households don't have access to a vehicle"

vehicle_dists <- data.frame(matrix(ncol = 1, nrow = 4))
colnames(vehicle_dists) = "Average distance to nearest supermarket (miles)"
vehicle_dists[1,1] = round(ave_dist_less_than_0.2 / 1609.344, digits = 2)
vehicle_dists[2,1] = round(ave_dist_less_than_0.4 / 1609.344, digits = 2)
vehicle_dists[3,1] = round(ave_dist_less_than_0.6 / 1609.344, digits = 2)
vehicle_dists[4,1] = round(ave_dist_greater_than_0.6 / 1609.344, digits = 2)

vehicle_table_combine <- cbind(vehicle_bracket,vehicle_dists)

vehicle_table <- kableExtra::kable(vehicle_table_combine)

kable_styling(vehicle_table, "striped", position = "left", font_size = 12, html_font = "arial")
```
  

# Conclusions

Results for the median household income data was a bit surprising. Given that supermarkets tend to favor more affluent neighborhoods, expectations were that as each income bracket increased, the distance would decrease. While the highest earners had to travel the least distance, the other three brackets did not show a linear progression of decreasing distances as income increased. However, the vehicle access data yielded expected results. As the proportion of households access to vehicles decreased, the average distance to supermarkets increased. This is expected because people without access to a vehicle typically come from vulnerable communities which supermarkets are less inclined to enter. 

Further research could be done by examining demographic variables such as race and education. By looking into these variables further insights could be gained to further explain the distribution of supermarkets in Buffalo, NY.

# References

- Crowe, J., Lacy, C., & Columbus, Y. (2018). Barriers to food security and community stress in an urban food desert. Urban Science, 2(2), 46. 

- Moore, L. V., & Diez Roux, A. V. (2006). Associations of neighborhood characteristics with the location and type of food stores. American journal of public health, 96(2), 325-331. 

- Sallis, J. F., Nader, P. R., Rupp, J. W., Atkins, C. J., & Wilson, W. C. (1986). San Diego surveyed for heart-healthy foods and exercise facilities. Public health reports (Washington, DC: 1974), 101(2), 216-219. 

- Walker, R. E., Keane, C. R., & Burke, J. G. (2010). Disparities and access to healthy food in the United States: A review of food deserts literature. Health & place, 16(5), 876-884. 
