---
title: "Supermarket presence by median household income and vehicle access in Buffalo, NY"
author: Eric Hughes
---

# Introduction 

Access to healthy food is not equal across all regions. Many factors influence food access such as vehicle access, proximity of quality grocery retailers, and income. The most basic measure of food access is proximity to supermarkets. Supermarkets have the resources to carry a wide range of produce and have affordable prices which makes their presence a good measure of local food security. The two main barriers to food are income level and access to a personal vehicle.

This project aims to look at Buffalo resident's proximity to supermarkets based off their median household income and access to a vehicle. To do this, two analysis will be run. For the first, Buffalo residents will be grouped into four different income brackets based on their median household income per census tract. For the second, Buffalo residents will be grouped into brackets based on proportions of households within each census tract without access to a personal vehicle. Then, the distance to the nearest supermarket for each census tract centroid will be calculated and. Each census tract will be joined to the median household income and vehicle access datasets and averages will be calculated for each bracket. Results will be analyzed and discussed in the Results and Conclusions sections. 

# Materials and methods

## Data sources

- US Census Bureau will provide census tract data of median household income, as well as demographic data, for Buffalo

- Open Data Buffalo for Neighborhood Boundary Shapefile

- ESRI Business Analyst will provide all supermarket data

## Packages
```{r, message=F, warning=F}
library(sf)
library(tidyverse)
library(ggplot2)
library(sp)
library(ggmap)
library(dodgr)
library(osmdata, quietly=T)
library(classInt)
library(geodist)
library(knitr)
```


## Data cleaning

```{r, message=F, warning=F}
#Create Variable with all census tracts within Buffalo boundary that contains median household income data

income_csv <- read.csv(file = "median_income_data_2020.csv",
                              skip = 2, header = F)
income_data <- income_csv[c(1,2,323)]
colnames(income_data) <- c("geoid", "census.tract", "median.household.income")
income_data$geoid = as.character(income_data$geoid)

vehicle_access_csv <- read.csv(file = "vehicle_access.csv")
vehicle_access_data <- vehicle_access_csv[c(1,2,6)]
colnames(vehicle_access_data) <- c("geoid","census.tract","prop.wo.va")

#Create Buffalo border

buffalo_border <- st_read("buffalo_boundary") %>%
  st_transform(4326)
buffalo_sp <- as_Spatial(st_transform(buffalo_border, 4326))
buffalo_bbox <- buffalo_sp@bbox

basemap <- get_stamenmap(
  bbox = buffalo_bbox,
  zoom = 13,
  maptype = "toner-lite")

#Import census tracts and join with median income and vehicle access data

census_tracts <- st_read("buffalo_census_tracts") 

income_shape <- census_tracts %>% left_join(income_data, by = c("GEOID" = "geoid"))
income_shape$INTPTLAT = as.numeric(income_shape$INTPTLAT)
income_shape$INTPTLON = as.numeric(income_shape$INTPTLON)
income_shape$median.household.income = as.integer(income_shape$median.household.income)

vehicle_shape <- census_tracts %>% left_join(vehicle_access_data, by = c("GEOID" = "geoid"))
vehicle_shape$INTPTLAT = as.numeric(vehicle_shape$INTPTLAT)
vehicle_shape$INTPTLON = as.numeric(vehicle_shape$INTPTLON)
vehicle_shape$prop.wo.va = as.double(vehicle_shape$prop.wo.va)

#Project data

income_proj <- income_shape %>% 
  st_transform(4326) 

vehicle_proj <- vehicle_shape %>% 
  st_transform(4326) 

#Turn census tracts into centroids

st_cent <- st_centroid(income_proj)

##Create Buffalo supermarkets variable##

#Create a 0.5-mile border around Buffalo
border_poly <- buffalo_border %>%
  st_cast("POLYGON")
buffer_buffalo <- st_buffer(border_poly, dist = 804.672)

#Select all the supermarkets within the new border
buffalo_supermarkets <- st_read("erie_county_supermarkets")%>%
  st_crop(buffer_buffalo)

#Create X and Y columns in buffalo_supermarkets and centroid data
coords = st_coordinates(buffalo_supermarkets)
xlon = coords[,1]
ylat = coords[,2]
bf_supermarkets <- buffalo_supermarkets %>%
  mutate(xlon,ylat) 

sf_coords <- st_as_sf(st_cent) 
cen_coords <- st_coordinates(sf_coords)
xlon = cen_coords[,1]
ylat = cen_coords[,2]

centroids <- sf_coords %>%
  mutate(xlon,ylat) %>%
  st_transform(4326)
centroids$id <- 1:max(nrow(centroids))

## Distance from centroid to nearest supermarket
# Define Buffalo coordinate variable
xy_coords <- rbind (
  c (-78.79514, 42.96641), # Buffalo
  c (-78.91252, 42.82610)
) # Buffalo
xy <- data.frame (lon = xy_coords [, 1], lat = xy_coords [, 2])

#Calculate distances from centroids to supermarkets

buffalo_network <- dodgr_streetnet (pts = xy, expand = 0.2, quiet = FALSE)
graph <- weight_streetnet (buffalo_network, wt_profile = "motorcar")
graph <- graph [which (graph$component == 1), ]
from_x <- centroids$INTPTLON
from_y <- centroids$INTPTLAT
to_x <- bf_supermarkets$xlon
to_y <- bf_supermarkets$ylat
d <- dodgr_dists(graph = graph, from = cbind (from_x, from_y), to = cbind (to_x, to_y))
df <- apply(d, 1, FUN=min, na.rm=TRUE) %>% as.data.frame()
b <- st_sf(data.frame(centroids, df))
names(b)[20] <- "dist"

##Create income brackets##

# Switch distance variable back to dataframe
b_df <- as_data_frame(b)

#Create brackets and join distance calculations
#Median income brackets
less_than_20k <- income_proj %>%
  filter(median.household.income <= 20000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_40k <- income_proj %>%
  filter(median.household.income > 20000, median.household.income <= 40000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_60k <- income_proj %>%
  filter(median.household.income > 40000, median.household.income <= 60000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

greater_than_or_equal_60k <- income_proj %>%
  filter(median.household.income >= 60000) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

#Vehicle access brackets
less_than_0.2 <- vehicle_proj %>%
  filter(prop.wo.va <= 0.2) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_0.4 <- vehicle_proj %>%
  filter(prop.wo.va > 0.2, prop.wo.va <= 0.4) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

less_than_0.6 <- vehicle_proj %>%
  filter(prop.wo.va > 0.4, prop.wo.va <= 0.6) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

greater_than_0.6 <- vehicle_proj %>%
  filter(prop.wo.va > 0.6) %>%
  left_join(b_df, by = c("GEOID" = "GEOID"))

##Calculate average distance for each bracket in meters##

ave_dist_less_than_20k <- mean(less_than_20k$dist) #1813.465
ave_dist_less_than_40k <- mean(less_than_40k$dist) #1926.935
ave_dist_less_than_60k <- mean(less_than_60k$dist) #1830.889
ave_dist_greater_than_60k <- mean(greater_than_or_equal_60k$dist) #1648.723

ave_dist_less_than_0.2 <- mean(less_than_0.2$dist) #1661.945
ave_dist_less_than_0.4 <- mean(less_than_0.4$dist) #1844.538
ave_dist_less_than_0.6 <- mean(less_than_0.6$dist) #1957.577
ave_dist_greater_than_0.6 <- mean(greater_than_0.6$dist) #2438.269
```

# Results

[~200 words]

Tables and figures (maps and other graphics) are carefully planned to convey the results of your analysis. Intense exploration and evidence of many trials and failures. The author looked at the data in many different ways before coming to the final presentation of the data.

Show tables, plots, etc. and describe them.

```{r, message=F, warning=F}
#Plot median income map with supermarkets

income_supermarket <- ggmap(basemap) +
  geom_sf(data = income_proj, aes(x = INTPTLON, y = INTPTLAT,  fill = median.household.income), 
          inherit.aes = F)+
  scale_fill_gradient(low = "yellow", high = "red") +
  geom_point(data = bf_supermarkets, aes(x = xlon, y = ylat, color = "Supermarket"), size = 3) +
  scale_color_manual(values = "green", name = "Supermarkets") +
  labs(title="All supermarkets within 0.5-miles of Buffalo, NY (2020)",
       subtitle="Median income data",
       caption="Median household income data from US Census Bureau,
       supermarket data from Data Axle",
       x = "Longitude",
       y = "Latitude") +
  theme(legend.direction = "vertical", axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1))+
  guides(fill=guide_legend(title="Median Household Income ($)"))
income_supermarket  

#Plot vehicle access data map with supermarkets

vehicle_supermarket <- ggmap(basemap) +
  geom_sf(data = vehicle_proj, aes(x = INTPTLON, y = INTPTLAT,  fill = prop.wo.va), 
          inherit.aes = F)+
  scale_fill_gradient(low = "#f1eef6", high = "#ce1256") +
  geom_point(data = bf_supermarkets, aes(x = xlon, y = ylat, color = "Supermarket"), size = 3) +
  scale_color_manual(values = "green", name = "Supermarkets") +
  labs(title="All supermarkets within 0.5-miles of Buffalo, NY (2020)",
       subtitle="Proportion of households without vehicle access",
       caption="Vehicle access data from US Census Bureau,
       supermarket data from Data Axle",
       x = "Longitude",
       y = "Latitude") +
  theme(legend.direction = "vertical", axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1))+
  guides(fill=guide_legend(title="Proportion of households 
without access to a vehicle")) 
vehicle_supermarket
  
```

More to come!
   

# Conclusions

[~200 words]

Coming soon!!

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.
